---
title: "Homework"
author: "Astrid Castro"
date: "5 November 2025"
embed-resources: true
format: html
---

The first two problems in this set refer to the `ISLR2::Carseats` set, which includes simulated sample data for 400 stores selling a particular brand of child carseat. We view this data as a simple random sample from a larger population.

```{r}
#| message: false
#| label: setup
#| warning: false

library(tidyverse)
library(ISLR2)
library(ggplot2)
```


## Problem 1

Does this data provide evidence against the company’s claim that the average price of their own seats is $110? Follow all best practices from class, including hypotheses, a relevant visualization, and a conclusion in non-technical language.

### Does the average price of the company's carseats = $110?

Null hypothesis: $\mu = 110$

Alternative hypothesis: $H_a: \mu \ne 110$

```{r}
#| fig-alt: A histogram of the distribution of the company's carseats' prices

ggplot(Carseats, aes(x = Price)) +
  geom_histogram(binwidth = 5,
                 fill = "skyblue", 
                 color = "black") +
  geom_vline(xintercept = 110,
             color = "red") +
  labs(title = "Distribution of Car Seat Prices",
       x = "Price ($)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```
```{r}
t.test(Carseats$Price, mu = 110)
```

### Conclusion: 

Based on the t-test, there is strong evidence that the average price of the company’s car seats is not $110. The sample suggests the true average price is higher,  115.795 to be exact. Besides, given the histogram of the distribution of prices for the company, we can see that we have a higher distribution of higher prices by looking at the red vertical line that shows the hypothesis average price. 

## Problem 2

Compute a level 90% confidence interval for the mean competitor price in the population from which this sample is drawn. What is the sample mean? What is the margin of error?

```{r}
t.test(Carseats$CompPrice, conf.level = 0.90)
```
Mean = 124.975

90% CI = [123.7109, 126.2391]

MOE = 1.2641


## Problem 3

Create a table with counts of penguin species and home islands in the penguins set. Include marginal counts in your table.

```{r}
table_species_island <- table(penguins$species, 
                              penguins$island)

table_with_margins <- addmargins(table_species_island)


table_with_margins

```


