---
title: "Homework"
author: "Astrid Castro"
date: "5 November 2025"
format: html
---

The first two problems in this set refer to the `ISLR2::Carseats` set, which includes simulated sample data for 400 stores selling a particular brand of child carseat. We view this data as a simple random sample from a larger population.

```{r}
#| message: false
#| label: setup
#| warning: false

library(tidyverse)
library(ISLR2)
library(ggplot2)
```


## Problem 1

Does this data provide evidence against the company’s claim that the average price of their own seats is $110? Follow all best practices from class, including hypotheses, a relevant visualization, and a conclusion in non-technical language.

### Does the average price of the company's seats = $110?

Null hypothesis: $\mu = 110$

Alternative hypothesis: $H_a: \mu \ne 110$

```{r}
t.test(Carseats$Price, mu = 110)
```

```{r}
ggplot(Carseats, aes(x = Price)) +
  geom_histogram(binwidth = 5,
                 fill = "skyblue", 
                 color = "black") +
  geom_vline(xintercept = 110,
             color = "red") +
  labs(title = "Distribution of Car Seat Prices",
       x = "Price ($)",
       y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```

Conclusion: Based on the t-test, there is strong evidence that the average price of the company’s car seats is not $110. The sample suggests the true average price is lower,  115.795 to be exact.

## Problem 2

Compute a level 90% confidence interval for the mean competitor price in the population from which this sample is drawn. What is the sample mean? What is the margin of error?

```{r}
t.test(Carseats$CompPrice, conf.level = 0.90)
```
Mean = 124.975

90% CI = [123.7109, 126.2391]

MOE = 1.2641


## Problem 3

Create a table with counts of penguin species and home islands in the penguins set. Include marginal counts in your table.

```{r}
table_species_island <- table(penguins$species, 
                              penguins$island)

table_with_margins <- addmargins(table_species_island)


table_with_margins

```


