---
title: "Early vs Late Respondents"
author: "Astrid Castro"
date: "24 October 2025"
embed-resources: true
format: html
---

## Introduction

The XKCD color survey had more than 200,000 participants taken over time in 2010.[^1] There was not a limit on how many questions each person got in the survey. Instead, the survey randomly generated colors to be named until the participant wanted to stop.[^2] Therefore, in our dataset, we see count responses per participant between a range of 1 to 1374. My hypothesis is: **Early respondents may have been more XKCD fans or people who like the topic of colours while later ones may have been more general internet users who saw the survey shared on blogs and social media.** This could imply that the sample composition shifted over time, potentially affecting how people labeled colors. I expect that early respondents will demonstrate greater engagement (higher number of responses per user) and more accurate or creative color naming, compared to later respondents. I am basing this on the assumption that the first people that completed the project where the interested, more familiar ones and the later people to complete it where people that tried it "for fun". 

### Indicators

To evaluate my hypothesis, I am interested in the variables `response`, `timestamp`, `user_id`, and `spam_prob`. I will take the first and last 25% of respondents to check the differences in **engagement, effort and creativity, and spam probability** between the groups. 

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(gt)
library(readr)
library(dplyr)
library(lubridate)
colors <- read_csv("color_survey.csv")
```

## Let's dive into it!

I will create a new dataframe with only the variables of interest. Then, I will use the `lubridate` library to fix the variable `timestamp`.[^3] This variable represents *Unix time* (the number of seconds since January 1, 1970). And since I want to use this variable to split the data into the two groups I will be comparing, I need to have a more readable information for it. Therefore, I will add a new column to our smaller dataset using the `as_datetime` argument. This new column will provide me the `timestamp` information in *2010-03-11 14:54:29 UTC* format.

```{r}
colors_sm <- colors %>% 
  select(user_id, timestamp, response, spam_prob) %>% 
  mutate(datetime = as_datetime(timestamp))
```

Let's visualize the response count during the time frame the survey went live.

```{r}
#| fig-alt: A histogram bar showing the distribution of responses during the timeframe the survery went live. 

ggplot(colors_sm, aes(x = datetime)) +
  geom_histogram(bins = 100,
                 binwidth = 86400, 
                 fill = "green") +
  labs(title = "Distribution of Response Times", 
       x = "Date", 
       y = "Responses Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```

Given this distribution, I see there are four areas where there are no responses at all. This could be because the survey closed or due to a glitch. I do not have enough information to make a conclusion about that. 

Moreover, I will split the data into two groups based on the `datetime` variable. For further information, I will make the `respondent_summary` subset that will have the response count per `user_id` and will be arranged by `datetime`. This is to later be able to compare engagement between groups.

```{r}
respondent_summary <- colors_sm %>%
  group_by(user_id) %>%
  summarise(response_count = n(),
            first_response_time = min(datetime, na.rm = TRUE)) %>%
  arrange(first_response_time)
```

Now let's use the `quantile` argument to get the cutoffs at the 25th and 75th percentile. [^4]

```{r}
quantile(respondent_summary$first_response_time, probs = c(0.25, 0.75))
```
Now let's create the two groups

```{r}
early_respondents <- respondent_summary %>% 
  filter(first_response_time <= "2010-03-03 00:47:13 UTC") %>% 
  mutate(group = "early")

late_respondents <- respondent_summary %>% 
  filter(first_response_time > "2010-03-20 08:07:01 UTC") %>% 
  mutate(group = "late")
```

Now I want to combine all the information I have gotten into my `respondent_summary` subset. The reason for this is because in that subset I already have a `response_count` per `user_id`, instead of duplicated `user_id`s in the `colors_sm` dataset. To do so, I will use the `bind_rows` and `left_join` functions.

```{r}
respondents <-
  bind_rows(early_respondents, late_respondents)

respondent_summary <- respondent_summary %>%
  left_join(respondents %>% 
              select(user_id, group),
            by = "user_id")

respondent_summary <- respondent_summary %>%
  left_join(colors_sm %>%
      group_by(user_id) %>%
      summarise(avg_spam_prob = mean(spam_prob, na.rm = TRUE),
                avg_response_length = mean(nchar(response), na.rm = TRUE)),
      by = "user_id")
```

For reference, let's do a quick summary table for `respondent_summary` but after filtering all `NA`s since I only want to include the `user_id`s that belong to our groups `early` and `late`. 

```{r}
respondent_summary_filtered <- respondent_summary %>%
  filter(!is.na(group))

summary(respondent_summary_filtered)
```
## Let's compare!

### Engagement

```{r}
#| fig-alt: A bar chart comparing the number of responses per user id between early and late respondents in the xkcd color survey.

ggplot(respondent_summary_filtered, aes(x = group, 
                               y = response_count, 
                               fill = group)) +
  geom_col(show.legend = FALSE, 
           position = "dodge") +
  labs(title = "Number of Responses per User ID",
       subtitle = "Early vs Late Respondents",
       y = "Response Count",
       x = "Group") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```

Late respondents exhibit a higher overall response count compared to early respondents, suggesting greater participation or continued engagement over time. After removing respondents with missing group classifications (NAs), this pattern indicates that those who responded later were more active within the survey or platform. This contrasts my hypothesis that early participants were the most engaged. 

### Creativity

```{r}
#| fig-alt: A bar chart comparing the average response length between early and late respondents in the xkcd color survey.

ggplot(respondent_summary_filtered, aes(x = group, 
                               y = avg_response_length, 
                               fill = group)) +
  geom_col(show.legend = FALSE, 
           position = "dodge") +
  labs(title = "Average Response Length per User ID",
  subtitle = "Early vs Late Respondents",
       y = "Average Characters",
       x = "Group") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```

Early respondents show longer average responses length compared to later participants. This supports the idea that those who found and completed the survey early were more engaged or interested in color naming, possibly being more detail-oriented users. In contrast, later respondents provided shorter answers on average, suggesting more casual participation. 

### Spam Probability

```{r}
#| fig-alt: A bar chart comparing the average spam probability between early and late respondents in the xkcd color survey.

respondent_summary_clean <- respondent_summary_filtered %>%
  filter(avg_spam_prob >= 0 & avg_spam_prob <= 1)

ggplot(respondent_summary_clean, aes(x = group, 
                               y = avg_spam_prob, 
                               fill = group)) +
  geom_col(show.legend = FALSE, 
           position = "dodge") +
  labs(title = "Average Spam Probability per User ID",
  subtitle = "Early vs Late Respondents",
       y = "Spam Probability",
       x = "Group") +
  theme_minimal() +
  scale_fill_brewer(palette = "Dark2")
```

Both groups display nearly identical averages, suggesting that the data quality was consistent throughout the survey’s duration. However, the values on `spam_prob` were not so accurate since many of them were above 1, which is invalid in probabilities. That's a limitation to come with a conclusion for this variable.

## Conclusion

Overall, the analysis supports the idea that early respondents were more knowledgeable with the xkcd color survey, as shown by their longer responses. This pattern aligns with the hypothesis that the first participants were likely more interested or familiar with the topic of color naming. However, while spam probability appeared consistent between groups, data irregularities in that variable limit any firm conclusions about response quality. Despite this, the findings highlight how participant behavior can evolve over time in open online surveys, shaping both the depth and reliability of collected data.

[^1]: [Color Survey Results](https://blog.xkcd.com/2010/05/03/color-survey-results/)
[^2]: [Color Name Survey](https://blog.xkcd.com/2010/03/01/color-name-survey/)
[^3]: [Equitable Equations](https://equitable-equations.quarto.pub/pdar_nsp/lubridate_new.html)
[^4]: [Quantile Function in R](https://www.digitalocean.com/community/tutorials/quantile-function-in-r)








